Select-based random access \\ to variable-byte encodings

### 2020-04-27 gradu/msc_english_contents.tex
Adds research questions to intro and expalantion to conclusions

In this thesis, an alternative solution that instead makes use of a $select$ data structure is proposed and explained in detail. Comparison to the $rank$-based 
implementation by Brisaboa et al is provided with different implementations of $rank$ and $select$ and several different data sets. The proposed method does not use data 
reorganizing, but rather capitalizes on the assumption that data is encoded the same way it is in the source. Because variable-byte encoding does not assign 
new codewords to data when encoding, data can be read straight from the memory and reassembling the integer from blocks is not needed.

Because of the block reorganizing, the $rank$ does not infact need a $rank$ query when fetching small elements. This makes the $rank$ method superior when dataset contains 
mostly small integers. With larger integers, the decoding needs multiple $rank$ calls and the new $select$-based method performs better.

Decompressing a subarray is experimented later in this thesis. Due to how the data is stored, the proposed $select$ method offers fast access to the next or previous element. This makes a 
subarray decompression very convenient and fast. Solutions for subarray fetching with both $rank$ and $select$ methods are proposed and compared. $Select$ based method performs better due to its 
single $select$ call requirement, while the $rank$ call amount depends on the size of the largest element in the subarray.

@@

It is not natively possible to decode the i-th element in variable-byte compression algorithms, because the position of the element in the compressed
list depends on the length of the preceding compressed data. Direct access is achievable with supporting data structures. A naive solution is to store the location 
of each element to an array, but it adds a very large overhead which removes the benefit of compression.

\section{Rank and Select}
$Rank$ and $select$ are two array operations that are widely used in compressed data structures. $Rank_x$($i$) counts the number of occurences of $x$ before index $i$ in the data. $Select_x$($i$) returns the index of the i-th 
occurence of $x$ in the data. In this thesis, these two operations are assumed to work with a bit array. $Rank_1$($i$) counts the number of bits set to 1 before $i$ and $select_1$($i$) returns the index 
of the i-th bit set to 1. Both operations can be implemented to work in constant time \citep[see, e.g.,][]{gbmp2014sea}. The two methods are related to each other: when used on a bit array 
$B$ = \texttt{0100 1101 \underline{1}011}, $select$(5) returns 8 (underlined) and therefore $rank$(8) returns 4.

@@

\section{DAC via rank}
\label{chap:implementations}
Directly addressable VB codes were first introduced by \citep{Bri09}. In their solution, the chunks are stored in separate arrays by their significance. For each integer, the block 
of the least significant chunk is stored in the first array $A_1$, and its continuation bit to the first bit array $B_1$. Then if the number was stored in multiple chunks, the block of the next chunk is
stored to $A_2$ and its continuation bit to $B_2$ and so on. After the data is set, a $rank$ data structure is built for each bit array. The data structure does not require additional space apart
from the $rank$ structure. Figure~\ref{bris_ds} contains a visualization of the data structure. The C array is the original VB encoded data, which is split into separate block arrays $A_k$ and 
contiuation bit arrays $B_k$.

@@

To decode an element from index $i$, first block is fetched from $A_1[i]$. Because each encoded element is composed of at least one block, the first block is obtainable via directly indexing $A_1$. 
With small integers, this is all that is required. If the data has a lot of small integers, the $rank$ method thus has a huge advantage. In addition, the block length of this method can change within 
different array levels which opens a possibility for further compression optimization.

@@

\begin{figure}[ht]
\centering
\begin{minipage}{0.5\linewidth}
\begin{algorithmic}
\State $index \gets $ wanted index
\State $A \gets $ block arrays
\State $B \gets $ continuation bit arrays
\State $level \gets 0$
\State $number \gets 0$
\While {$B[level][index] = 0$}
\State $block \gets A[level][index]$
\State $number \gets number \mathbin{\ll} 8$
\State $number \gets number + block$
\State $index \gets \Call{Rank}{B[level], index}$
\State $level \gets level + 1$
\EndWhile
\State $block \gets A[level][index]$
\State $number \gets number \mathbin{\ll} 8$
\State $number \gets number + block$
\end{algorithmic}
\end{minipage}
\caption{Example pseudo code of DAC with rank by Brisaboa et al.} \label{bris_pseudo}

@@

Currently $P$ is [0,0,-1,-1]. The location of the first block of the next element is fetched by increasing $P$[0] by 1 and then using the incremented value. The bit array shows that this element is only 
one block long. The first two blocks of the third element is fetched the same way. $P$[2] is unset, so $P$[2] = $rank_0$($B_2$,0) is used for the third block, which is the last block of that element. After 
these elements, $P$ is [2,1,0,-1]. The index locations of the fourth element are fetched from $P$ the same way. In the end $P$ is [3,2,0,-1] and the process could potentially go on indefinitely with only a 
very small amount of $rank$ calls.

@@

@@

The order of how data is stored with $select$ compression enables an efficient fetching of a subarray. Only one $select$ query is needed in the beginning, and the decoding of the elements past the first one 
consists of just a few bit operations.

There has already been a few successful attempts to decompress VB encoded integers with SIMD instructions \citep{Lem18,Pla15}. It is worthwhile to check if the select-based approach presented 
in this thesis benefits from SIMD. The data needs less manipulation because the data is already stored in the correct order, but calculating the integer block lengths might benefit from SIMD.

@@

 - list research questions somewhere? (introduction 1.1?)

@@

7. 
 - write a recap about what was done

>>>>

In this thesis, an alternative solution that instead makes use of a $select$ data structure is proposed and explained in detail. The proposed method does not use data 
reorganizing, but rather capitalizes on the assumption that data is encoded the same way it is in the source. Because variable-byte encoding does not assign 
new codewords to data when encoding, data can be read straight from the memory and reassembling the integer from blocks is not needed.

Comparison of the proposed method and $rank$-based implementation by Brisaboa et al is provided with different implementations of $rank$ and $select$. The $rank$ method 
is assumed to perform better on data sets containing mostly small integers. It is assumed that larger numbers cause problems in performance. The performance of the proposed $select$ method is indifferent 
of the size of the elements in the data. Different kinds of data sets are used to see how each algorithm performs. The assumption is that $select$ works better with data sets with very large integers. It
remains to be seen how large the integers need to be for the $select$ method to outperform the $rank$ method.

Another interesting point is to compare the algorithms when decompressing a long subarray. Due to how the data is stored, the proposed $select$ method offers fast access to the next or previous element 
and is expected to perform well with any given data set or subarray length. Solutions for subarray fetching with both $rank$ and $select$ methods are proposed and compared.

@@

\chapter{Rank and Select}
$Rank$ and $select$ are two array operations that are widely used in compressed data structures. $Rank_x$($i$) counts the number of occurences of $x$ before index $i$ in the data. $Select_x$($i$) returns the index of the i-th 
occurence of $x$ in the data. In this thesis, these two operations are assumed to work with a bit array. $Rank_1$($i$) counts the number of bits set to 1 before $i$ and $select_1$($i$) returns the index 
of the i-th bit set to 1. Both operations can be implemented to work in constant time \citep[see, e.g.,][]{gbmp2014sea}. The two operations are related to each other: when used on a bit array 
$B$ = \texttt{0100 1101 \underline{1}011}, $select$(5) returns 8 (underlined) and therefore $rank$(8) returns 4.

@@

It is not natively possible to decode the $i$-th element in variable-byte compression algorithms, because the position of the element in the compressed
list depends on the length of the preceding compressed data. Direct access is achievable with supporting data structures. A naive solution is to store the location 
of each element to an array, but it adds a very large overhead which removes the benefit of compression.

\section{DAC via rank}
\label{chap:implementations}
Directly addressable VB codes were first introduced by \citep{Bri09}. In their solution, the chunks are stored in separate arrays by their significance. For each integer, the block 
of the least significant chunk is stored in the first array $A_1$, and its continuation bit to the first bit array $B_1$. Then if the number was stored in multiple chunks, the block of the next chunk is
stored to $A_2$ and its continuation bit to $B_2$ and so on.

After the data is set, a $rank$ data structure is built for each bit array. The data structure does not require additional space apart
from the $rank$ structure. Figure~\ref{bris_ds} contains a visualization of the data structure. The C array is the original VB encoded data, which is split into separate block arrays $A_k$ and 
contiuation bit arrays $B_k$.

@@

To decode an element from index $i$, first block is fetched from $A_1[i]$. Because each encoded element is composed of at least one block, the first block is obtainable via directly indexing $A_1$. 
With small integers, this is all that is required. If the data has a lot of small integers, the $rank$ method thus has a huge advantage.

\pagebreak

@@

Interestingly the block length of this method can change within different array levels which opens a possibility for further compression optimization. \citep{Bri09} experiments a few different block 
length setups and leaves open the problem of finding optimal block lengths. While the problem of finding the optimal block length for each data set in regular VB encoding is interesting, the possibility to
optimize it for each block level separately is surely more challenging.
\begin{figure}[ht]
\centering
\begin{minipage}{0.5\linewidth}
\begin{algorithmic}
\State $index \gets $ wanted index
\State $A \gets $ block arrays
\State $B \gets $ continuation bit arrays
\State $level \gets 0$
\State $number \gets 0$
\While {$B[level][index] = 0$}
\State $block \gets A[level][index]$
\State $number \gets number \mathbin{\ll} 8$
\State $number \gets number + block$
\State $index \gets \Call{Rank}{B[level], index}$
\State $level \gets level + 1$
\EndWhile
\State $block \gets A[level][index]$
\State $number \gets number \mathbin{\ll} 8$
\State $number \gets number + block$
\end{algorithmic}
\end{minipage}
\caption{Example pseudo code of DAC with rank by Brisaboa et al.} \label{bris_pseudo}

@@

Currently $P$ is [0,0,-1,-1]. The location of the first block of the next element is fetched by increasing $P$[0] by 1 and then using the incremented value. The bit array shows that this element is only 
one block long. The first two blocks of the third element is fetched the same way. $P$[2] is unset, so $P$[2] = $rank_0$($B_2$,0) is used for the third block, which is the last block of that element. After 
these elements, $P$ is [2,1,0,-1]. The index locations of the fourth element are fetched from $P$ the same way. In the end $P$ is [3,2,0,-1] and the process could continue if needed.

@@

This thesis explained variable-byte encoding, a compression method that stores the data in fixed length blocks and omits empty leading blocks. A continuation bit is added to each block to signal where an 
element ends and a new one begins. $Rank$ and $select$, the two bit array operations were also explained. Existing solution for direct access with VB using $rank$ was explained. A new solution for direct 
access with $select$ was proposed and results were compared between the two solutions. A method of extracting an array of data with VB encoding was proposed for both $rank$ and $select$.

@@

The order of how data is stored with $select$ compression enables an efficient fetching of a subarray. Only one $select$ query is needed in the beginning, and the decoding of the elements past the first one 
consists of just a few bit operations. The amount of $rank$ queries depend on the length of the subarray and the average size of the elements. Therefore even a small amount of large elements in an array 
otherwise filled with small elements causes the runtime to go up. This leads to the conclusion that the $select$ method, if not strictly superior, is a strong contender and a good option for decompressing 
a subarray.

\section{Future work}
There has already been a few successful attempts to decompress VB encoded integers with SIMD instructions \citep{Lem18,Pla15}. It is worthwhile to check if the select-based approach presented 
in this thesis benefits from SIMD. The data needs less manipulation because the data is already stored in the correct order, but calculating the integer block lengths might benefit from SIMD.

@@

@@

 - write more in conclusions
